---
title: "LLM-Human Reliability Report - `r params$run_label`"
author: "Generated report"
date: "`r Sys.Date()`"
params:
  run_label: "unnamed run"
  predictions_path: ""
  raw_responses_path: ""
  split_path: ""
  human_reference_path: ""
  audit_csv_path: ""
  inspected_ids_path: ""
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
    theme: flatly
---

```{r setup, include=FALSE}
# Parameterized version of the v1 report
# (v1/results/2026-06-09_llm-human-irr_dev-train_gemma4-31b_limit-20.Rmd).
# The analysis is identical; all run-specific paths come in via params so the
# same template scores every version. Render through
# scripts/render_irr_report.R rather than knitting directly.
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 100
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)
library(tibble)
library(knitr)
library(scales)
library(jsonlite)
library(readr)

prediction_path <- params$predictions_path
raw_response_path <- params$raw_responses_path
split_path <- params$split_path
human_reference_path <- params$human_reference_path
audit_csv_path <- params$audit_csv_path

read_jsonl <- function(path) {
  jsonlite::stream_in(file(path), verbose = FALSE)
}

cohen_kappa <- function(tab) {
  tab <- as.matrix(tab)
  if (nrow(tab) != ncol(tab)) return(NA_real_)
  n <- sum(tab)
  if (n == 0) return(NA_real_)
  po <- sum(diag(tab)) / n
  pe <- sum(rowSums(tab) * colSums(tab)) / (n^2)
  if (pe == 1) return(NA_real_)
  (po - pe) / (1 - pe)
}

yes_flag <- function(x) {
  coalesce(str_to_lower(str_trim(as.character(x))) == "yes", FALSE)
}

label_bloom <- function(x) {
  x <- as.character(x)
  case_when(
    is.na(x) ~ NA_character_,
    x %in% c("Nonpossession", "Nonposession") ~ "nonexistence",
    x == "Nonexistence" ~ "nonexistence",
    x == "Rejection" ~ "rejection",
    x == "Denial" ~ "denial",
    x == "Uncoded" ~ "uncoded",
    x == "Excluded" ~ "excluded",
    TRUE ~ "other"
  )
}

kappa_fixed <- function(a, b, levels) {
  cohen_kappa(table(factor(a, levels), factor(b, levels)))
}

context_to_text <- function(x) {
  if (is.null(x) || length(x) == 0 || all(is.na(x))) return(NA_character_)
  if (!is.data.frame(x)) return(as.character(x))
  x %>%
    transmute(text = paste0(line, " ", speaker, " ", utterance)) %>%
    pull(text) %>%
    paste(collapse = " | ")
}

flag_cols <- c(
  "not_a_negation",
  "foreign_language_negation",
  "mimicry",
  "singing",
  "tag_question",
  "repetition"
)

label_levels <- c("nonexistence", "rejection", "denial", "uncoded", "excluded", "other")
content_label_levels <- c("nonexistence", "rejection", "denial")
```

# Overview

This report evaluates a Bloom LLM coding run as an additional independent
coder, following the validation plan in
`Data/XLing-LLM_coding/LLM_validation_plan.md`.

```{r run-metadata}
run_meta <- tibble(field = "Predictions file", value = basename(prediction_path))
if (nzchar(raw_response_path) && file.exists(raw_response_path)) {
  raw_meta <- read_jsonl(raw_response_path)
  batch_sizes <- vapply(raw_meta$record_ids, length, integer(1))
  run_meta <- bind_rows(
    run_meta,
    tibble(
      field = c("Model", "Schema version", "Prompt version", "Run date", "Batches (records per batch)"),
      value = c(
        as.character(raw_meta$model[[1]]),
        as.character(raw_meta$schema_version[[1]]),
        as.character(raw_meta$prompt_version[[1]]),
        substr(as.character(raw_meta$raw_response$created_at[[1]]), 1, 10),
        paste0(length(batch_sizes), " (", paste(unique(batch_sizes), collapse = ", "), ")")
      )
    )
  )
}
run_meta <- bind_rows(
  run_meta,
  tibble(
    field = c("Split file", "Human reference file"),
    value = c(split_path, human_reference_path)
  )
)
run_meta %>% kable(caption = "Run snapshot.")
```

The denominator rules match the human-human IRR reports:

* Blank human coder rows are missing data, not `not-a-negation` judgments.
* Bloom agreement requires both sides to have non-missing Bloom labels.
* `Nonpossession` is collapsed to `nonexistence`.
* `Uncoded` and `Excluded` remain explicit labels.
* For flags, only explicit `Yes` counts as `TRUE`.

```{r load-data}
predictions_raw <- read_jsonl(prediction_path)
split_raw <- read_jsonl(split_path)
human_raw <- read_jsonl(human_reference_path)

predictions <- predictions_raw %>%
  transmute(
    record_id,
    llm_bloom_raw = bloom_label,
    llm_label = label_bloom(bloom_label),
    llm_comments = comments,
    llm_not_a_negation = yes_flag(flags$not_a_negation),
    llm_foreign_language_negation = yes_flag(flags$foreign_language_negation),
    llm_mimicry = yes_flag(flags$mimicry),
    llm_singing = yes_flag(flags$singing),
    llm_tag_question = yes_flag(flags$tag_question),
    llm_repetition = yes_flag(flags$repetition)
  )

human <- human_raw %>%
  transmute(
    record_id,
    source_row,
    coder_1_sheet,
    coder_2_sheet,
    coder_1_name = coder_1$coded_by_raw,
    coder_2_name = coder_2$coded_by_raw,
    human_1_bloom_raw = coder_1$bloom_label,
    human_2_bloom_raw = coder_2$bloom_label,
    human_1_label = label_bloom(coder_1$bloom_label),
    human_2_label = label_bloom(coder_2$bloom_label),
    human_1_comments = coder_1$comments,
    human_2_comments = coder_2$comments,
    human_1_not_a_negation = yes_flag(coder_1$flags$not_a_negation),
    human_2_not_a_negation = yes_flag(coder_2$flags$not_a_negation),
    human_1_foreign_language_negation = yes_flag(coder_1$flags$foreign_language_negation),
    human_2_foreign_language_negation = yes_flag(coder_2$flags$foreign_language_negation),
    human_1_mimicry = yes_flag(coder_1$flags$mimicry),
    human_2_mimicry = yes_flag(coder_2$flags$mimicry),
    human_1_singing = yes_flag(coder_1$flags$singing),
    human_2_singing = yes_flag(coder_2$flags$singing),
    human_1_tag_question = yes_flag(coder_1$flags$tag_question),
    human_2_tag_question = yes_flag(coder_2$flags$tag_question),
    human_1_repetition = yes_flag(coder_1$flags$repetition),
    human_2_repetition = yes_flag(coder_2$flags$repetition)
  )

split_info <- split_raw %>%
  transmute(
    record_id,
    split,
    language,
    transcript_id,
    half,
    line,
    speaker,
    child_id,
    child_age_months,
    child_age_raw,
    target_negator,
    target_utterance,
    context_before = map_chr(context_before, context_to_text),
    context_after = map_chr(context_after, context_to_text),
    coded_by_1,
    coded_by_2
  )

eval_rows <- predictions %>%
  left_join(split_info, by = "record_id") %>%
  left_join(human, by = "record_id") %>%
  mutate(
    has_human_1 = !is.na(human_1_label),
    has_human_2 = !is.na(human_2_label),
    has_llm = !is.na(llm_label),
    human_agreed = has_human_1 & has_human_2 & human_1_label == human_2_label,
    human_disagreed = has_human_1 & has_human_2 & human_1_label != human_2_label
  )

long_labels <- bind_rows(
  eval_rows %>% transmute(record_id, coder = "human_1", coder_label = coder_1_sheet, label = human_1_label),
  eval_rows %>% transmute(record_id, coder = "human_2", coder_label = coder_2_sheet, label = human_2_label),
  eval_rows %>% transmute(record_id, coder = "llm", coder_label = "LLM", label = llm_label)
)

pair_data <- list(
  `human_1 vs human_2` = c("human_1", "human_2"),
  `human_1 vs LLM` = c("human_1", "llm"),
  `human_2 vs LLM` = c("human_2", "llm")
) %>%
  imap_dfr(function(pair, pair_name) {
    long_labels %>%
      filter(coder %in% pair) %>%
      select(record_id, coder, label) %>%
      pivot_wider(names_from = coder, values_from = label) %>%
      transmute(
        record_id,
        pair = pair_name,
        a = .data[[pair[[1]]]],
        b = .data[[pair[[2]]]]
      )
  })

# Inspected rows are development data (read, mined for prompt examples, or
# adjudicated); headline metrics are computed on the remaining rows.
inspected_ids <- character(0)
if (nzchar(params$inspected_ids_path) && file.exists(params$inspected_ids_path)) {
  inspected_lines <- readLines(params$inspected_ids_path, warn = FALSE)
  inspected_lines <- trimws(inspected_lines)
  inspected_ids <- inspected_lines[nzchar(inspected_lines) & !startsWith(inspected_lines, "#")]
}
eval_rows <- eval_rows %>% mutate(inspected = record_id %in% inspected_ids)
pair_data <- pair_data %>% mutate(inspected = record_id %in% inspected_ids)
```

# Coverage

```{r coverage}
coverage <- tibble(
  measure = c(
    "Predicted rows",
    "Rows with human coder 1 Bloom label",
    "Rows with human coder 2 Bloom label",
    "Rows with both human Bloom labels",
    "Rows with at least one human Bloom label",
    "Rows with LLM plus at least one human Bloom label"
  ),
  n = c(
    nrow(eval_rows),
    sum(eval_rows$has_human_1),
    sum(eval_rows$has_human_2),
    sum(eval_rows$has_human_1 & eval_rows$has_human_2),
    sum(eval_rows$has_human_1 | eval_rows$has_human_2),
    sum(eval_rows$has_llm & (eval_rows$has_human_1 | eval_rows$has_human_2))
  )
)

coverage %>%
  kable(caption = "Coverage for the predicted subset.")
```

# Bloom Agreement

```{r bloom-agreement}
pair_summary <- pair_data %>%
  filter(!is.na(a), !is.na(b)) %>%
  group_by(pair) %>%
  summarise(
    n_denominator = n(),
    pct_agreement = mean(a == b),
    kappa = kappa_fixed(a, b, label_levels),
    denial_agreement = mean((a == "denial") == (b == "denial")),
    denial_kappa = kappa_fixed(a == "denial", b == "denial", c(FALSE, TRUE)),
    excluded_agreement = mean((a == "excluded") == (b == "excluded")),
    excluded_kappa = kappa_fixed(a == "excluded", b == "excluded", c(FALSE, TRUE)),
    uncoded_agreement = mean((a == "uncoded") == (b == "uncoded")),
    uncoded_kappa = kappa_fixed(a == "uncoded", b == "uncoded", c(FALSE, TRUE)),
    .groups = "drop"
  )

pair_summary %>%
  mutate(
    across(ends_with("agreement"), ~ percent(.x, accuracy = 0.1)),
    across(ends_with("kappa"), ~ round(.x, 3))
  ) %>%
  kable(caption = "Bloom-label agreement. Denominator is rows where both compared coders have non-missing Bloom labels.")
```

## Headline vs Inspected Rows

```{r inspected-stratification, results='asis'}
if (!length(inspected_ids)) {
  cat("No inspected-rows file was supplied; all rows are treated as uninspected.\n")
} else {
  n_inspected_here <- sum(eval_rows$inspected)
  cat(sprintf(
    paste0(
      "%d of the %d predicted rows are on the inspected list ",
      "(`%s`): their content informed prompt examples or adjudications, so ",
      "they are development data. **Headline numbers are the clean-subset ",
      "rows**; the inspected rows are a rule-compliance check only.\n\n"
    ),
    n_inspected_here, nrow(eval_rows), params$inspected_ids_path
  ))

  strat_summary <- pair_data %>%
    filter(!is.na(a), !is.na(b)) %>%
    mutate(subset = if_else(inspected, "inspected (compliance check)", "clean (headline)")) %>%
    group_by(subset, pair) %>%
    summarise(
      n_denominator = n(),
      pct_agreement = mean(a == b),
      kappa = kappa_fixed(a, b, label_levels),
      .groups = "drop"
    )

  if (nrow(strat_summary)) {
    print(
      strat_summary %>%
        mutate(
          pct_agreement = percent(pct_agreement, accuracy = 0.1),
          kappa = round(kappa, 3)
        ) %>%
        arrange(desc(subset == "clean (headline)"), pair) %>%
        kable(caption = "Bloom-label agreement stratified by inspection status.")
    )
  }

  consensus_strat <- eval_rows %>%
    filter(human_agreed, has_llm) %>%
    mutate(subset = if_else(inspected, "inspected (compliance check)", "clean (headline)")) %>%
    group_by(subset) %>%
    summarise(
      n_denominator = n(),
      pct_agreement = mean(llm_label == human_1_label),
      .groups = "drop"
    )

  if (nrow(consensus_strat)) {
    print(
      consensus_strat %>%
        mutate(pct_agreement = percent(pct_agreement, accuracy = 0.1)) %>%
        arrange(desc(subset == "clean (headline)")) %>%
        kable(caption = "LLM agreement with the agreed-human label, stratified by inspection status.")
    )
  }
}
```

## Contentful Bloom Agreement Only

```{r content-agreement}
content_summary <- pair_data %>%
  filter(a %in% content_label_levels, b %in% content_label_levels) %>%
  group_by(pair) %>%
  summarise(
    n_denominator = n(),
    pct_agreement = mean(a == b),
    kappa = kappa_fixed(a, b, content_label_levels),
    .groups = "drop"
  )

content_summary %>%
  mutate(
    pct_agreement = percent(pct_agreement, accuracy = 0.1),
    kappa = round(kappa, 3)
  ) %>%
  kable(caption = "Content-label agreement only: nonexistence, rejection, and denial.")
```

## Confusion Tables

```{r confusion-tables, results='asis'}
for (pair_name in unique(pair_data$pair)) {
  cat("\n\n### ", pair_name, "\n\n", sep = "")
  pair_df <- pair_data %>%
    filter(pair == pair_name, !is.na(a), !is.na(b))
  print(
    table(
      `coder A` = factor(pair_df$a, label_levels),
      `coder B` = factor(pair_df$b, label_levels)
    ) %>%
      as.data.frame.matrix() %>%
      knitr::kable(caption = paste("Confusion table:", pair_name))
  )
}
```

# Agreed-Human Subset

```{r agreed-human-subset}
agreed_subset <- eval_rows %>%
  filter(human_agreed, has_llm) %>%
  mutate(agreed_human_label = human_1_label)

agreed_summary <- agreed_subset %>%
  summarise(
    n_denominator = n(),
    pct_agreement = mean(llm_label == agreed_human_label),
    kappa = kappa_fixed(llm_label, agreed_human_label, label_levels)
  )

agreed_summary %>%
  mutate(
    pct_agreement = percent(pct_agreement, accuracy = 0.1),
    kappa = round(kappa, 3)
  ) %>%
  kable(caption = "LLM agreement with the consensus label on rows where both humans agree.")
```

# Human-Disagreement Siding

```{r disagreement-siding}
disagreement_siding <- eval_rows %>%
  filter(human_disagreed, has_llm) %>%
  mutate(
    llm_position = case_when(
      llm_label == human_1_label ~ "agrees with human 1",
      llm_label == human_2_label ~ "agrees with human 2",
      llm_label %in% c("uncoded", "excluded") ~ paste("LLM used", llm_label),
      TRUE ~ "agrees with neither"
    )
  ) %>%
  count(llm_position, name = "n")

if (nrow(disagreement_siding) == 0) {
  tibble(note = "No rows in this predicted subset where both humans coded and disagreed.") %>%
    kable(caption = "Human-disagreement siding.")
} else {
  disagreement_siding %>%
    mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
    kable(caption = "How the LLM behaves on rows where both humans disagree.")
}
```

# Flag Agreement

```{r flag-agreement}
flag_pair_summary <- map_dfr(flag_cols, function(flag) {
  pairs <- list(
    `human_1 vs human_2` = c(paste0("human_1_", flag), paste0("human_2_", flag)),
    `human_1 vs LLM` = c(paste0("human_1_", flag), paste0("llm_", flag)),
    `human_2 vs LLM` = c(paste0("human_2_", flag), paste0("llm_", flag))
  )

  imap_dfr(pairs, function(cols, pair_name) {
    df <- eval_rows %>%
      filter(!is.na(.data[[cols[[1]]]]), !is.na(.data[[cols[[2]]]]))
    tibble(
      flag = flag,
      pair = pair_name,
      n_denominator = nrow(df),
      pct_agreement = if_else(nrow(df) > 0, mean(df[[cols[[1]]]] == df[[cols[[2]]]]), NA_real_),
      kappa = if_else(nrow(df) > 0, kappa_fixed(df[[cols[[1]]]], df[[cols[[2]]]], c(FALSE, TRUE)), NA_real_),
      n_yes_a = sum(df[[cols[[1]]]], na.rm = TRUE),
      n_yes_b = sum(df[[cols[[2]]]], na.rm = TRUE),
      n_both_yes = sum(df[[cols[[1]]]] & df[[cols[[2]]]], na.rm = TRUE),
      n_one_yes = sum(xor(df[[cols[[1]]]], df[[cols[[2]]]]), na.rm = TRUE)
    )
  })
})

flag_pair_summary %>%
  mutate(
    pct_agreement = percent(pct_agreement, accuracy = 0.1),
    kappa = round(kappa, 3)
  ) %>%
  arrange(flag, pair) %>%
  kable(caption = "Flag agreement. Only explicit Yes values are TRUE; No, blanks, and NA are FALSE.")
```

# Item-Level Audit

The CSV written next to this report contains every child negation predicted in
this run, with context, model coding, and both human codings:

`r audit_csv_path`

```{r audit-table}
audit_table <- eval_rows %>%
  transmute(
    record_id,
    inspected,
    split,
    language,
    transcript_id,
    half,
    line,
    speaker,
    target_negator,
    target_utterance,
    context_before,
    context_after,
    llm_bloom = llm_bloom_raw,
    llm_label_collapsed = llm_label,
    llm_comments,
    llm_foreign_language_negation,
    llm_singing,
    llm_mimicry,
    llm_tag_question,
    llm_repetition,
    llm_not_a_negation,
    human_1_sheet = coder_1_sheet,
    human_1_coder = coder_1_name,
    human_1_bloom = human_1_bloom_raw,
    human_1_label_collapsed = human_1_label,
    human_1_comments,
    human_1_foreign_language_negation,
    human_1_singing,
    human_1_mimicry,
    human_1_tag_question,
    human_1_repetition,
    human_1_not_a_negation,
    human_2_sheet = coder_2_sheet,
    human_2_coder = coder_2_name,
    human_2_bloom = human_2_bloom_raw,
    human_2_label_collapsed = human_2_label,
    human_2_comments,
    human_2_foreign_language_negation,
    human_2_singing,
    human_2_mimicry,
    human_2_tag_question,
    human_2_repetition,
    human_2_not_a_negation
  )

readr::write_csv(audit_table, audit_csv_path, na = "")

audit_table %>%
  select(
    record_id,
    transcript_id,
    line,
    target_negator,
    target_utterance,
    llm_bloom,
    human_1_sheet,
    human_1_bloom,
    human_2_sheet,
    human_2_bloom,
    llm_comments
  ) %>%
  kable(caption = "Compact item-level audit table. See the CSV for full before/after context and all flags.")
```

# Notes

```{r notes, results='asis'}
n_rows <- nrow(eval_rows)
n_single_human <- sum(xor(eval_rows$has_human_1, eval_rows$has_human_2))
cat(sprintf(
  paste0(
    "This run predicts %d rows%s, so kappa values should be treated as ",
    "descriptive diagnostics rather than stable performance estimates when n ",
    "is small. %d row%s in this subset have only one human Bloom label; those ",
    "contribute to LLM-vs-human comparisons where applicable but not to the ",
    "human-human baseline.\n\n"
  ),
  n_rows,
  if (n_rows < 100) " (a smoke-test-sized subset)" else "",
  n_single_human,
  if (n_single_human == 1) "" else "s"
))
cat("After rendering this report, rebuild the cross-version viewer with `Rscript scripts/build_coding_viewer.R` so the run appears in the run selector and on the agreement-across-versions chart.\n")
```
